Mapping mutational effects along the evolutionary landscape of HIV envelope

### 2017-12-12 paper/manuscript_Sarah.tex
started adding Sarah comments

\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}

@@

\begin{abstract}
The immediate evolutionary space accessible to HIV is largely determined by how single amino-acid mutations affect fitness.
These mutational effects can shift as the virus evolves.
However, the prevalence of such shifts remains unclear.
Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
Most mutations similarly affect both Envs, but the amino-acid preferences of a minority of sites have clearly shifted.
These shifted sites usually prefer a specific amino acid in one Env, but tolerate many amino acids in the other.
Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs -- and many occur at residues that do not even contact substitutions.
Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the amino-acid preferences of most sites are conserved between moderately diverged viral strains.
\end{abstract}

@@

Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
Most sites show no detectable shifts, but 30 sites have clearly altered their preferences.
These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
The shifted sites cluster in structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
Overall, our work provide complete across-strain maps of mutational effects that inform the analysis of Env's evolution.

@@

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{nduati2000effect,wu2006neutralization,goo2014early}.
The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
We used the T332N variant of BG505 Env because it has a common glycosylation site that is targeted by many anti-HIV antibodies~\citep{sanders2013next}.
The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.34\textwidth}
{\bf \Large A} \\ 
\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
\centerline{\small BG505}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
\centerline{\small BF520}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
\end{minipage}
\begin{minipage}[t]{0.01\textwidth}
\end{minipage}
\begin{minipage}[t]{0.65\textwidth}
{\bf \Large B} \\
\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
\end{minipage}
\caption{\label{fig:mutfreqs}
The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
{\bf (A)}
The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
{\bf (B)}
Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
$R$ is the Pearson correlation coefficient. 
The numerical values for the preferences are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:allprefs}.
}
\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in Figure~\ref{fig:mutfreqs}.
Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
{\input{figures/avgmutfreqs}}
\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
\end{figure}

The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
Of course, Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data.
Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset. 
Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.

@@

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1.
All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

>>>>

\newcommand{\jdbcomment}[1]{\emph{\color{red} [#1]}}
\newcommand{\skhcomment}[1]{\emph{\color{magenta} [#1]}}

@@

\begin{abstract}
The immediate evolutionary space accessible to HIV is largely determined by how single amino-acid mutations affect fitness.
These mutational effects can shift as the virus evolves.
However, the prevalence of such shifts remains unclear.
Here we quantify the effects on viral replication of all amino-acid mutations to two HIV envelope (Env) proteins that differ at $>$100 residues.
Most mutations similarly affect both Envs, but the amino-acid preferences of a minority of sites have clearly shifted.
These shifted sites usually prefer a specific amino acid in one Env, but tolerate many amino acids in the other.
Surprisingly, shifts are only slightly enriched at sites that have substituted between the Envs -- and many occur at residues that do not even contact substitutions.
Therefore, long-range epistasis can unpredictably shift Env's mutational tolerance during HIV evolution, although the amino-acid preferences of most sites are conserved between moderately diverged viral strains.
\end{abstract}
\skhcomment{How ubiquitous is the term "shifted"? }

@@

Here we use an improved version of a previously described deep mutational scanning strategy~\citep{haddox2016experimental} to measure the effects on viral replication of all single amino-acid mutations to two transmitted-founder virus Envs that differ by $>$100 mutations.
We compare these complete maps of mutational effects to identify sites that have shifted in their amino-acid preferences between the Envs.
Most sites show no detectable shifts, but 30 sites have clearly altered preferences.
These shifted sites usually prefer a specific amino acid in one Env but have shifted to tolerate many amino acids in the other Env.
The shifted sites cluster in structure, but are often distant from any amino-acid substitutions that distinguish the two Envs, demonstrating the action of long-range epistasis.
By aggregating our measurements for both Envs, we identify sites that evolve faster or slower in nature than expected given the functional constraints measured in the lab, probably due to pressure for immune evasion.
Overall, our work provide complete across-strain maps of mutational effects that inform the analysis of Env's evolution.

@@

\begin{figure}
\centerline{\includegraphics[clip=true, trim= 0in 0in 1in 0in, angle=-90, width=0.85\textwidth]{figures/tree_plot.pdf}}
\caption{\label{fig:tree}
Phylogenetic tree showing the relationship of BG505 and BF520 to other clade A Envs.
The tree shows the 69 Envs in the alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, which is a subsample of clade A sequences from the group M alignment in the Los Alamos HIV sequence database (\url{http://www.hiv.lanl.gov}).
Sites not mutagenized in our experiments (the signal peptide and cytoplasmic tail) or that are poorly alignable (mostly sites in the variable loops) were masked as indicated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}, leaving 616 alignable sites.
The pairwise identity of BG505 and BF520 to the other sequences at these alignable sites is shown in Figure~\ref{fig:tree}-Figure~supplement~\ref{figsupp:identity}.
The tree topology was inferred using \texttt{RAxML}~\citep{stamatakis2014raxml} under the GTRCAT model of nucleotide substitution, and branch lengths were optimized under the M0 version of the Goldman-Yang model~\citep{yang2000codon} using \texttt{phydms}~\citep{hilton2017phydms}.
\skhcomment{Maybe change the color of the BG505 and BF520 labels so they stick out more? Also, add a label for some sort of landmark sequence (such as the oldest sequence) to contextualize the tree.}
}
\figsupp[\label{figsupp:identity}  Pairwise identity of all Env sequences in our clade A alignment to BG505 and BF520 after masking poorly alignable or non-mutagenized sites.]
{The histograms show the pairwise amino-acid identity of each Env to all other sequences in the clade A alignment in Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment} after masking the sites delineated in Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}.
There are 616 non-masked sites.
The pairwise protein identity between BG505 and BF520 is 86.2\% (721 of 836 sites identical) when considering \emph{all} sites, and 89.1\% (549 of 616 sites identical) when considering just the non-masked sites.}
{\includegraphics[width=0.7\textwidth]{figures/masked_alignment_identity.pdf}}
\figdata{\label{figdata:alignment} The alignment of clade A \textit{env} coding sequences is in \texttt{cladeA\_alignment.fasta}.}
\figdata{\label{figdata:mask} The 240 Env sites masked in all phylogenetic analyses because they were not mutagenized in our experiments or are poorly alignable are listed in \texttt{alignment\_mask.csv}.}
\end{figure}

We therefore selected Envs from two transmitted-founder viruses, BG505.W6M.C2.T332N and BF520.W14M.C2 (hereafter referred to as BG505 and BF520), that were isolated from HIV-infected infants shortly after mother-to-child transmission~\citep{goo2014early}.
The BG505 Env has been extensively studied from a structural standpoint~\citep{julien2013crystal,lyumkis2013cryo,pancera2014structure,huang2014broad,sanders2015hiv,stewart2016trimeric}, and variants of this Env are being tested as a vaccine immunogens~\citep{sanders2013next,sanders2015hiv,de2015immunogenicity}.
We used the T332N variant of BG505 Env because it has a common glycosylation site that is targeted by many anti-HIV antibodies~\citep{sanders2013next}.
The BF520 Env was isolated from an infant who developed an early broad anti-HIV antibody response~\citep{goo2014early,simonich2016hiv}.
We have previously created comprehensive codon-mutant libraries of the BF520 Env and used them to map HIV antibody escape~\citep{dingens2017comprehensive}, but these BF520 libraries have not been characterized with respect to how mutations affect viral replication.

Both BG505 and BF520 are from clade A of the major (M) group of HIV-1.
Figure~\ref{fig:tree} shows the phylogenetic relationship among these two Envs and other clade A sequences.
BG505 and BF520 have typical divergence \skhcomment{Do you mean they have roughly the mean divergence? I might change the wording or provide this value.} for two clade A Envs: they are identical at 721 of the 836 pairwise-alignable protein sites (86.2\% identity), and 549 of the 616 sites (89.1\% identity) in the ectodomain and transmembrane domain that are alignable across clade A (Figure~\ref{fig:tree}-source~data~\ref{figdata:alignment}, Figure~\ref{fig:tree}-source~data~\ref{figdata:mask}).
\skhcomment{I am not sure I understand the difference between the denominator 836 and the denominator 616.}
The divergence between BG505 and BF520 therefore offers ample opportunity to investigate mutational shifts during Env evolution.

@@

\begin{figure}
\centerline{\includegraphics[width=0.75\textwidth]{figures/dms_schematic/dms_schematic.pdf}}
\caption{\label{fig:dms_schematic}
Deep mutational scanning workflow.
{\bf (A)} We made libraries of proviral HIV plasmids with random codon-level mutations in the \textit{env} gene.
The number of mutations per gene followed a roughly Poisson distribution with a mean between 1 and 1.5 (Figure~\ref{fig:dms_schematic}-Figure~supplement~\ref{figsupp:sanger}).
We transfected the plasmids into 293T cells to generate mutant viruses, which lack a genotype-phenotype link since cells are multiply transfected.
To establish a genotype-phenotype link and select for Env variants that support HIV replication, we passaged the libraries in SupT1.CCR5 cells for four days at a low multiplicity of infection (MOI) of 0.01.
To isolate the \textit{env} genes from only viruses that encoded a functional Env protein, we infected the passaged libraries into SupT1.CCR5 cells at high MOI and harvested reverse-transcribed non-integrated viral DNA after 12 hours.
We then deep sequenced the \textit{env} genes from these final samples as well as the initial plasmid library, using molecular barcoding to reduce sequencing errors.
We also deep sequenced identically handled wildtype controls to estimate error rates.
Using these sequencing data, we estimated the preference for each of the 20 amino acids at each site in Env.
These data are represented in logo plots, with the height of each letter proportional to that site's preference for that amino acid.
{\bf (B)} We conducted this experiment in full biological triplicate for both BG505 and BF520, beginning each replicate with independent creation of the plasmid mutant library.
These replicates therefore account for \emph{all} sources of noise and error in the experiments.
\skhcomment{Three comments on the schematic. 1. The genome strands might be too wide for the size of the figure. 2. I think the grey plasmid should be changed to orange. 3. The dark and light orange are difficult to tell apart.}
}
\figsupp[Sanger sequencing of selected clones from the mutant plasmid libraries.\label{figsupp:sanger}]{
We Sanger sequenced 44 clones of BG505 Env sampled roughly evenly from each of the three replicate mutant plasmid libraries.
{\bf(A)} There was an average of 1.5 mutant codons per clone, with the number of mutations per clone roughly following a Poisson distribution. 
{\bf(B)} The mutant codons had a mix of single-, double-, and triple-nucleotide changes, with an elevated number of single-nucleotide changes than expected.
{\bf(C)} Nucleotide frequencies were fairly uniform in the mutant codons.
{\bf(D)} Mutations were distributed roughly evenly along the mutagenized region of {\it env} (30-699 in the sequential numbering scheme used in this plot).
{\bf(E)} For clones with multiple mutations, we computed the pairwise distance in primary sequence between each codon mutation and plotted the cumulative distribution of these distances (red line). 
We also simulated the expected distribution of pairwise distances if mutations occurred entirely independently (blue line). 
The observed distribution is close to the expected distribution.
Comparable data for the BF520 libraries is provided in \citet{dingens2017comprehensive}.
}{\includegraphics[width=\textwidth]{figures/sanger_sequencing_supp/sanger_sequencing_supp.pdf}}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.34\textwidth}
{\bf \Large A} \\ 
\centerline{{\includegraphics[clip=true, trim=4.3in 1.4in 0in 0.48in, width=0.4\textwidth]{figures/BG505_avgmutfreqs.pdf}}} 
\centerline{\small BG505}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=\textwidth]{figures/BG505_avgmutfreqs.pdf}
\centerline{\small BF520}
\includegraphics[clip=true, trim=0in 0in 1.7in 0in, width=0.89\textwidth]{figures/BF520_avgmutfreqs.pdf}
\end{minipage}
\begin{minipage}[t]{0.01\textwidth}
\end{minipage}
\begin{minipage}[t]{0.65\textwidth}
{\bf \Large B} \\
\includegraphics[width=\textwidth]{figures/allprefscorr.pdf}
\end{minipage}
\caption{\label{fig:mutfreqs}
The deep mutational scanning selects for functional Envs and yields measurements that are well correlated among replicates.
{\bf (A)}
The average per-codon mutation frequency when sequencing plasmids encoding wildtype Env (\emph{DNA}), plasmid mutant libraries (\emph{mutDNA}), mutant viruses after the final infection (\emph{mutvirus}), and virus generated from wildtype plasmids (\emph{virus}).
Mutations are categorized as nonsynonymous, synonymous, or stop codon. 
The \emph{DNA} samples show that sequencing errors are rare, and the \emph{virus} samples show that viral replication and reverse-transcription errors are well below the frequency of mutations in the \emph{mutDNA} samples.
Comparing the \emph{mutvirus} to \emph{mutDNA} shows clear purifying selection against stop codons and some nonsynonymous mutations, particularly after subtracting the background error rates given by the \emph{virus} and \emph{DNA} samples (Figure~\ref{fig:mutfreqs}-Figure~supplement~\ref{figsupp:mutfreqs}).
More extensive plots from the analysis of the deep sequencing data are in Supplementary~files~\ref{suppfile:code} and \ref{suppfile:html}.
{\bf (B)}
Correlations between replicates in the measured preferences of each site in Env for all 20 amino acids.
Blue indicates replicate measurements on BF520, red indicates replicate measurements on BG505, and gray indicates across-Env measurements of BF520 versus BG505.
$R$ is the Pearson correlation coefficient. 
The numerical values for the preferences are in Figure~\ref{fig:mutfreqs}-source~data~\ref{figdata:allprefs}.
\skhcomment{1. The axis text is a little bit blurry on my copy. 2. I would swap the order of barplots so that BF520 is close to the BF520 correlation plots. 3. Could the mutation type legend be horizontal?}
}
\figsupp[\label{figsupp:mutfreqs} Numerical frequencies of mutations.]
{This table gives the average frequencies of nonsynonymous, synonymous, and stop-codon mutations as plotted in Figure~\ref{fig:mutfreqs}.
Note that there is only one \emph{DNA} sample for BF520, so that same sample is repeated three times in the sample in conjunction with each BF520 replicate.
We calculate the error-corrected \emph{pre}-selection mutation frequency as the \emph{mutDNA} frequency minus the \emph{DNA} frequency, and the error-corrected \emph{post}-selection mutation frequency as the \emph{mutvirus} frequency minus the \emph{virus} frequency.
We then use these error-corrected frequencies to calculate the \emph{percent} of mutations remaining after selection for each replicate and type of mutation.}
{\input{figures/avgmutfreqs}}
\figdata{\label{figdata:allprefs} Preferences for each replicate and averages are in \texttt{all\_prefs\_unscaled.zip}.}
\end{figure}

The deep mutational scanning effectively selected for functional Env as evidenced by strong purifying selection against stop codons.
Figure~\ref{fig:mutfreqs}A shows the average frequency of mutations across Env in the plasmid mutant libraries, the mutant viruses, and wildtype controls as determined from the deep sequencing.
The mutant viruses show clear selection against stop codons and many nonsynonymous mutations (Figure~\ref{fig:mutfreqs}A).
This selection is more apparent if we correct for the background error rates estimated from the wildtype controls (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}).
\skhcomment{Could you discuss selection only in the context of error-corrected rates?}
The error-corrected frequencies of stop codons drop to 3\%--16\% of their original values (Figure~\ref{fig:mutfreqs}-Figure-supplement~\ref{figsupp:mutfreqs}), with the residual stop codons probably due to some non-functional virions surviving due to complementation by other co-infecting virions. 
The error-corrected frequencies of nonsynonymous mutations also drop substantially (43\%--49\% of their original values), whereas the frequencies of synonymous mutations drop only slightly (85\%--95\% of their original values).
These trends are consistent with the fact that nonsynonymous mutations are often deleterious, whereas synonymous mutations often~\citep[although certainly not always, see][]{zanini2013quantifying} have only mild effects on viral replication.
Figure~\ref{fig:mutfreqs}A only summarizes one aspect of the deep mutational scanning data but supplementary~files~\ref{suppfile:code} and \ref{suppfile:html} contain detailed plots showing all aspects of the data (read depth, per-site mutation rate, etc) as generated by the \texttt{dms\_tools2} software~\citep[\url{https://jbloomlab.github.io/dms_tools2/}]{bloom2015software}.

We used the deep mutational scanning data to estimate the preference of each site in Env for each amino acid via the analysis method described in \citet{bloom2015software}.
As graphically illustrated in Figure~\ref{fig:dms_schematic}A, the preferences for each site are normalized to sum to one.
Our libraries were mutagenized at 670 sites in BG505 and 662 sites in BF520, so $670 \times 20 = 13,400$ and $662 \times 20 = 13,240$ preferences were estimated for each Env, respectively.
The correlations between the preferences from different experimental replicates are in Figure~\ref{fig:mutfreqs}B.
These replicate-to-replicate correlations are substantially higher than those for the deep mutational scanning of LAI Env by \citet{haddox2016experimental}, indicating that the experimental modifications used here were beneficial.
Because of the clear superiority of the current measurements, in this paper we do not attempt to compare to our earlier inferior LAI dataset. 
Figure~\ref{fig:mutfreqs}B also shows that while replicates are well correlated across BG505 and BF520, the replicates for BG505 are more correlated with each other than with replicates for BF520, and vice versa.
\skhcomment{Later on in the text this claim is more conservative and takes into account the fact that BF520 is basically as well correlated with itself as with BG505.}
This fact hints that there are some shifts in amino-acid preferences between the two Envs -- something that is investigated with more statistical rigor later in this paper.

@@

\begin{table}
\begin{fullwidth}
{\centering \include{figures/gamma_omega_modelcomparison}}
\caption{\label{tab:phydms}
Evolutionary models informed by the deep mutational scanning describe HIV's evolution in nature much better than a standard substitution model.
Shown are the results of maximum likelihood fitting of substitution models to the clade A phylogeny in Figure~\ref{fig:tree}.
Experimentally informed codon models~\citep[ExpCM,][]{hilton2017phydms} utilizing the across-replicate average of the deep mutational scanning describe Env's natural evolution far better than a standard codon substitution model~\citep[the M5 model of][]{yang2000codon} as judged by comparing the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}.
Both ExpCM models have a stringency parameter $>$1, indicating selection in nature prefers the same amino acids as the experiments, but with greater stringency.
All models draw $\omega$ from a gamma distribution, and the table shows the mean ($\overline{\omega}$) and shape parameters ($\omega_{\alpha}$ and $\omega_{\beta}$) of this distribution.
The last two columns show the number of sites evolving faster ($\omega_r > 1$) or slower ($\omega_r < 1$) than expected at a false discovery rate of 0.05, as determined using the approach in \citet{bloom2017identification}.
Analyses were performed using \texttt{phydms} (\url{http://jbloomlab.github.io/phydms/}).
Table~\ref{tab:phydms}-source-data~\ref{tabledata:phydms} shows the results for additional substitution models.
}
\tabledata{\label{tabledata:phydms}Results for phylogenetic models where $\omega$ is not drawn from a gamma-distribution or where the preferences are averaged across sites to eliminate the site specificity are in \texttt{modelcomparison.md}.}
\end{fullwidth}
\end{table}

Table~\ref{tab:phydms} shows that phylogenetic models informed by the deep mutational scanning of either BG505 or BF520 describe the natural evolution of Env vastly better than a standard codon substitution model.
In addition to simply comparing model fits, we can interpret the model parameters.
ExpCM's have a stringency parameter that relates selection in the experiments to that in nature.
A stringency parameter $>$1 indicates that natural selection prefers the same amino acids as the experiments, but with greater stringency~\citep{hilton2017phydms}.
Both ExpCM's have stringency parameters $>$1 (Table~\ref{tab:phydms}) -- a finding that makes sense, since the stop-codon analysis in the previous section suggests that the experimental selections are more lax than natural selection on HIV.
\skhcomment{Comment here that re-scaled preferences are used throughout the rest of the study?}
We can also interpret the $\omega$ parameter.
For standard codon substitution models, $\omega$ is the rate of fixation of nonsynonymous mutations relative to synonymous one -- and the gene-wide average $\omega$ is almost always $<$1, since purifying selection purges many functionally deleterious amino-acid mutations even for adaptively evolving proteins~\citep{murrell2015gene}.
Indeed, Table~\ref{tab:phydms} shows that Env's gene-wide average $\omega$ is $<$1 for a standard model.
\skhcomment{Is it common to report the mean $\omega$ when a model uses $\Gamma\omega$? I usually think of "gene-wide average $\omega$" to be $\omega$ from M0.}
But for ExpCM, $\omega$ is the relative rate of nonsynonymous to synonymous substitutions \emph{after} accounting for functional constraints measured in the deep mutational scanning~\citep{bloom2017identification}.
Table~\ref{tab:phydms} shows that the ExpCM's have a gene-wide average $\omega$ that is $>$1, indicating that immune selection drives Env to fix amino-acid mutations faster than expected under a null model that only accounts for functional constraints.
